On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-05 manuscripts/PNAS/chazthm_pnas.tex
main paper SRed

\begin{lemma}\label{MainLemma}
%Fix integers $k, \ell, m, \bar m$ and $n$ with $k < m$. 
Suppose $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} for some $k < m$ and that $E \subseteq {[m] \choose k}$ is $\ell$-regular and satisfies the SIP. There is a constant $C_2 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_2$.

@@

Given a sequence of vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$, we let $\mathbf{X}$ be the $m \times N$ matrix with columns $\mathbf{x}_i$ and $I(S) := \{i \in [N]: \text{supp}(\mathbf{x}_i) \subseteq S\}$. The constant $C_1$ in Thm.~\ref{DeterministicUniquenessTheorem} is then given by\footnote{Note that $\|\mathbf{AX}_{I(S)}\mathbf{c}\|_2 \geq \sqrt{k} L_k(\mathbf{A})\|\mathbf{X}_{I(S)}\mathbf{c}\|_2 \geq k L_k(\mathbf{A}) L_k(\mathbf{X}_{I(S)})\|\mathbf{c}\|_2$ for $S \in E$ and $k$-sparse $\mathbf{c}$. Therefore, $\sqrt{k} L_k(\mathbf{AX}_{I(S)}) \geq k L_k(\mathbf{A}) L_k(\mathbf{X}_{I(S)}) > 0$ since $L_k(\mathbf{A}), L_k(\mathbf{X}_{I(S)}) > 0$ by \eqref{SparkCondition} and general linear position of the $\mathbf{x}_i$.  Thus, $C_1 > 0$.  }:
\begin{align}\label{Cdef1}
C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E) := \frac{ C_2(\mathbf{A}, E) } { \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) }.
\end{align}

The constant $C_2 = C_2(\mathbf{A}, E)$ is defined below in terms of one used in \cite{Deutsch12} to analyze the convergence of the alternating projections algorithm.
% for projecting a point onto the intersection of a set of subspaces. 
We use it to bound the distance between a point and the intersection of a set of subspaces given an upper bound on its distance from each subspace individually.

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For subspaces $V_1, \ldots, V_\ell \subseteq \mathbb{R}^m$, set $r := 1$ when $\ell = 1$ and define for $\ell \geq 2$:
\begin{align*}
r(\{V_i\}_{i=1}^\ell) := 1 - \left(1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right)  \right)^{1/2},
\end{align*} 
%
where the maximum is taken over all orderings\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < r \leq 1$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U = V$ and when $U \perp V$.}  %\textbf{Note:} we can also define $\theta$ by \cite[Lem.~9.5]{Deutsch12}:
%\begin{align*}
%\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \mathbf{u} \in U \cap (U \cap W)^\perp, \|\mathbf{u}\|_2 \leq 1, \mathbf{w} \in W \|\mathbf{w}\|_2 \leq 1  \right\}.
%\end{align*}
The constant $C_2$ in Lem.~\ref{MainLemma} can then be expressed as:  % \footnote{Note that $C_2 > 0$ is well-defined since $r > 0$ by definition.}
\begin{align}\label{Cdef2}
C_2(\mathbf{A}, E) := \frac{ 2^{|E|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{F \subseteq E} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F}) }.
\end{align}
%Note that when $k=1$ the constant $C_1$ is consistent with the value in our proof of the case $k=1$.

\begin{proof}[Proof of Thm.~\ref{DeterministicUniquenessCorollary} for $k < m$] 
%We begin by showing that for every $S \in E$ there is some $S' \in {[m] \choose k}$ for which the distance $d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_S'\})$ is controlled by $\varepsilon$. Fix $S \in E$. Since there are $(k-1){m \choose k}+1$ vectors $\mathbf{x}_i$ supported on $S$, the pigeon-hole principle implies that there exists some $S' \in {[m] \choose k}$ and some set of $k$ indices $J(S)$ such that the supports of all $\mathbf{x}_i$ and $\mathbf{\bar x}_i$ with $i \in J(S)$ are contained in $S$ and $S'$, respectively.
%%% No |S| = k assumption %%%
We shall show that for every $S \in E$ there is some $\bar S \in {[m] \choose k}$, for which the distance $d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\bar S}\})$ is controlled by $\varepsilon$. Applying Lem.~\ref{MainLemma} with the map $\pi$ defined by $S \mapsto \bar S$ then completes the proof.  Since there are $(k-1){\bar m \choose k}+1$ vectors $\mathbf{x}_i$ supported on $S$, the pigeon-hole principle implies that there exists some $\bar S \in {[\bar m] \choose k}$ and some set of $k$ indices $K$ such that the supports of all $\mathbf{x}_i$ and $\mathbf{\bar x}_i$ with $i \in K$ are contained in $S$ and $\bar S$, respectively.

%Let $\mathbf{X}$ and $\mathbf{\bar{X}}$ be the $m \times N$ matrices with columns $\mathbf{x}_i$ and $\mathbf{\bar x}_i$, respectively. 
It follows from the general linear position of the $\mathbf{x}_i$ and the linear independence of every $k$ columns of $\mathbf{A}$ that $L_k(\mathbf{AX}_{I(S)}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\text{\rm Span}\{\mathbf{A}_{S}\}$. Fixing $\mathbf{y} \in \text{\rm Span}\{\mathbf{A}_{S}\}$, there exists $\mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\bar{y}} = \mathbf{B\bar{X}}_K\mathbf{c} \in \text{\rm Span}\{\mathbf{B}_{\bar S}\}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\bar{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\bar{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
%\leq \frac{\varepsilon \sqrt{k}}{L(\mathbf{A}X_{J(S)})} \|\mathbf{A}X_{J(S)}\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L_k(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality is by definition of $L_k$. Finally, from Def.~\ref{dDef}, it follows that for all $S \in E$,
\begin{align}\label{rhs222}
d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\bar S}\}) 
\leq \frac{\varepsilon}{ \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) }.
\end{align}
\end{proof}

@@

We will show that no more than $(k-1)|E|$ vectors $\mathbf{\bar x}_i$ can share a support of cardinality less than $k$. 
%Let $\mathbf{X}$ and $\mathbf{\bar{X}}$ be the $m \times N$ and $\bar m \times N$ matrices with columns $\mathbf{x}_i$ and $\mathbf{\bar x}_i$, respectively, and let $I(S) = \{i \in [N]: \text{supp}(\mathbf{x}_i) \subseteq S\}$. 
From previous arguments, we have for every $S \in E$, the columns of every $n \times k$ submatrix of $\mathbf{AX}_{I(S)}$ form a basis for $\text{\rm Span}\{\mathbf{A}_S\}$.
% the general linear position of the $\mathbf{x}_i$ and the linear independence of every $k$ columns of $\mathbf{A}$ that $L_k(\mathbf{AX}_{I(S)}) > 0$ for all $S \in E$; that is, for every $S \in E$, the columns of every $n \times k$ submatrix of $\mathbf{AX}_{I(S)}$ form a basis for $\text{\rm Span}\{\mathbf{A}_S\}$. 
Suppose there are more than $(k-1)|E|$ vectors $\mathbf{\bar x}_i$ sharing a support $\bar S$ of cardinality less than $k$. By the pigeonhole principle, there must be some $S \in E$ supporting $k$ or more of the corresponding $\mathbf{x}_i$; let them be indexed by $K \subseteq I(S)$.  By the argument as in the proof of Thm.\ref{DeterministicUniquenessTheorem}, we also have \eqref{rhs222}. Since our bound on $\varepsilon$ implies that the right-hand side of \eqref{rhs222} is less than one, by \eqref{dimLem} we have $k = \dim(\text{Span}(\mathbf{A}_S)) \leq \dim(\text{Span}\{\mathbf{B}_{\bar S}\}) \leq |\bar S|$, a contradiction.

Hence, the number of $(k-1)$-sparse vectors $\mathbf{\bar x}_i$ is no greater than $|E|(k-1){ m \choose k-1}$. Taking \eqref{eqn} into account, it follows that at least $N - |E|k(k-1){ m \choose k-1}$ of the vectors $\mathbf{\bar  x}_i$ are $k$-sparse. Therefore, more than $(k-1){m \choose k}$ of the vectors $\mathbf{x}_i$ supported on each $S \in E$ have corresponding $\mathbf{\hat x}_i$ that are also $k$-sparse. The result now follows directly from Thm.~\ref{DeterministicUniquenessCorollary}.
\end{proof}

>>>>

\begin{lemma}\label{MainLemma}
%Suppose $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} for some $k < m$ and that $E \subseteq {[m] \choose k}$ is $\ell$-regular and satisfies the SIP. There is a constant $C_2 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_2$. 
Suppose $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} for some $k < m$ and that $E \subseteq {[m] \choose k}$ is $\ell$-regular with the SIP. There exists $C_2 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_2$.

@@

Given vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$, we let $\mathbf{X}$ be the $m \times N$ matrix with columns $\mathbf{x}_i$ and $I(S) := \{i \in [N]: \text{supp}(\mathbf{x}_i) \subseteq S\}$. The constant $C_1$ in Thm.~\ref{DeterministicUniquenessTheorem} is then defined by\footnote{Note that $\|\mathbf{AX}_{I(S)}\mathbf{c}\|_2 \geq \sqrt{k} L_k(\mathbf{A})\|\mathbf{X}_{I(S)}\mathbf{c}\|_2 \geq k L_k(\mathbf{A}) L_k(\mathbf{X}_{I(S)})\|\mathbf{c}\|_2$ for $S \in E$ and $k$-sparse $\mathbf{c}$. Therefore, $\sqrt{k} L_k(\mathbf{AX}_{I(S)}) \geq k L_k(\mathbf{A}) L_k(\mathbf{X}_{I(S)}) > 0$ since $L_k(\mathbf{A}), L_k(\mathbf{X}_{I(S)}) > 0$ by \eqref{SparkCondition} and general linear position of the $\mathbf{x}_i$.  Thus, $C_1 > 0$.  }:
\begin{align}\label{Cdef1}
C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E) := \frac{ C_2(\mathbf{A}, E) } { \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) }.
\end{align}

The constant $C_2 = C_2(\mathbf{A}, E)$ is given below in terms of one used in \cite{Deutsch12} to analyze the convergence of the alternating projections algorithm.
% for projecting a point onto the intersection of a set of subspaces. 
We use it to bound the distance between a point and the intersection of a set of subspaces given an upper bound on its distance from each subspace individually.

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For subspaces $V_1, \ldots, V_\ell \subseteq \mathbb{R}^m$, set $r := 1$ when $\ell = 1$ and define for $\ell \geq 2$:
\begin{align*}
r(\{V_i\}_{i=1}^\ell) := 1 - \left(1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right)  \right)^{1/2},
\end{align*} 
%
where the maximum is taken over all orderings\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < r \leq 1$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U = V$ or $U \perp V$.}  %\textbf{Note:} we can also define $\theta$ by \cite[Lem.~9.5]{Deutsch12}:
%\begin{align*}
%\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \mathbf{u} \in U \cap (U \cap W)^\perp, \|\mathbf{u}\|_2 \leq 1, \mathbf{w} \in W \|\mathbf{w}\|_2 \leq 1  \right\}.
%\end{align*}
The constant $C_2$ in Lem.~\ref{MainLemma} can then be expressed as:  % \footnote{Note that $C_2 > 0$ is well-defined since $r > 0$ by definition.}
\begin{align}\label{Cdef2}
C_2(\mathbf{A}, E) := \frac{ 2^{|E|} \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{F \subseteq E} r( \{ \text{\rm Span}\{\mathbf{A}_{S}\} \}_{S \in F}) }.
\end{align}
Note that $k=1$ gives the same $C_1$ as that used to prove $k=1$.

\begin{proof}[Proof of Thm.~\ref{DeterministicUniquenessCorollary} for $k < m$] 
We shall show that for every $S \in E$ there is some $\bar S \in {[\bar m] \choose k}$, for which the distance $d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\bar S}\})$ is controlled by $\varepsilon$. Applying Lem.~\ref{MainLemma} with the map $\pi$ defined by $S \mapsto \bar S$ then completes the proof.

Since there are $(k-1){\bar m \choose k}+1$ vectors $\mathbf{x}_i$ supported on $S$, the pigeonhole principle implies that there is some $\bar S \in {[\bar m] \choose k}$ and some set of $k$ indices $K$ such that the supports of all $\mathbf{x}_i$ and $\mathbf{\bar x}_i$ with $i \in K$ are contained in $S$ and $\bar S$, respectively.

%Let $\mathbf{X}$ and $\mathbf{\bar{X}}$ be the $m \times N$ matrices with columns $\mathbf{x}_i$ and $\mathbf{\bar x}_i$, respectively. 
It follows from the general linear position of the $\mathbf{x}_i$ and the linear independence of every $k$ columns of $\mathbf{A}$ that $L_k(\mathbf{AX}_{K}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\text{\rm Span}\{\mathbf{A}_{S}\}$. Fixing $\mathbf{0} \neq \mathbf{y} \in \text{\rm Span}\{\mathbf{A}_{S}\}$, there exists $\mathbf{0} \neq \mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\bar{y}} = \mathbf{B\bar{X}}_K\mathbf{c} \in \text{\rm Span}\{\mathbf{B}_{\bar S}\}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\bar{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\bar{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
%\leq \frac{\varepsilon \sqrt{k}}{L(\mathbf{A}X_{J(S)})} \|\mathbf{A}X_{J(S)}\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L_k(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality follows directly from the definition of $L_k$. Finally, from Def.~\ref{dDef}, it follows that:
\begin{align}\label{rhs222}
d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\bar S}\}) 
\leq \frac{\varepsilon}{  L_k(\mathbf{AX}_{K}) }.
\end{align}
In particular, with $C_1$ as in \eqref{Cdef1}, the theorem follows.
\end{proof}
% for all $S \in E$:
%\begin{align}\label{rhs222}
%d(\text{\rm Span}\{\mathbf{A}_S\}, \text{\rm Span}\{\mathbf{B}_{\bar S}\}) 
%\leq \frac{\varepsilon}{ \min_{S \in E} L_k(\mathbf{AX}_{I(S)}) }.
%\end{align}
%\end{proof}

@@

We now show that no more than $(k-1)|E|$ of the $\mathbf{\bar x}_i$ share a support of size less than $k$. 
%Let $\mathbf{X}$ and $\mathbf{\bar{X}}$ be the $m \times N$ and $\bar m \times N$ matrices with columns $\mathbf{x}_i$ and $\mathbf{\bar x}_i$, respectively, and let $I(S) = \{i \in [N]: \text{supp}(\mathbf{x}_i) \subseteq S\}$. 
From previous arguments, for every $S \in E$, the columns of each $n \times k$ submatrix of $\mathbf{AX}_{I(S)}$ form a basis for $\text{\rm Span}\{\mathbf{A}_S\}$.
% the general linear position of the $\mathbf{x}_i$ and the linear independence of every $k$ columns of $\mathbf{A}$ that $L_k(\mathbf{AX}_{I(S)}) > 0$ for all $S \in E$; that is, for every $S \in E$, the columns of every $n \times k$ submatrix of $\mathbf{AX}_{I(S)}$ form a basis for $\text{\rm Span}\{\mathbf{A}_S\}$. 
Suppose that more than $(k-1)|E|$ vectors $\mathbf{\bar x}_i$ share a support $\bar S$ of size $|\bar{S}| < k$. By the pigeonhole principle, there is some $S \in E$ supporting $k$ or more of the corresponding $\mathbf{x}_i$; let them be indexed by $K \subseteq I(S)$.  By the argument as in the proof of Thm.\ref{DeterministicUniquenessTheorem}, we also have \eqref{rhs222}. Since our bound on $\varepsilon$ implies that the right-hand side of \eqref{rhs222} is less than one, by \eqref{dimLem}: \[k = \dim(\text{Span}(\mathbf{A}_S)) \leq \dim(\text{Span}\{\mathbf{B}_{\bar S}\}) \leq |\bar S|,\] a contradiction.

Hence, the number of $(k-1)$-sparse vectors $\mathbf{\bar x}_i$ is no greater than $|E|(k-1){ m \choose k-1}$. Taking \eqref{eqn} into account, it follows that at least $N - |E|k(k-1){ m \choose k-1}$ of the vectors $\mathbf{\bar  x}_i$ are $k$-sparse. Therefore, more than $(k-1){m \choose k}$ of the vectors $\mathbf{x}_i$ supported on each $S \in E$ have corresponding $\mathbf{\bar x}_i$ that are also $k$-sparse. The result now follows directly from Thm.~\ref{DeterministicUniquenessCorollary}.
\end{proof}
